---
title: "K-Means Analysis"
author: "Gloria Wu"
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
library(ggplot2)
library(tidyverse)
library(GGally)

# Load datasets
load("data/k-means-data.Rds")
```

## K-Means Analysis

To conduct the $k$-means analysis, only numerical variables were included, and they were all standardized. Then, an elbow plot was constructed using the ggplot2 package by @ggplot2 to see the optimal amount of clusters to conduct the $k$-means analysis. Finally, visualizations for the $k$-means analysis were generated using the tidyverse package by @tidyverse and GGally package by @GGally. 

### Elbow Plot

```{r}
#| label: elbow plot for k-means analysis

elbow_plot |>
  ggplot(aes(x = k, y = tot.withinss)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Number of clusters (k)",
       y = "Total within-cluster sum of squares")
```

The elbow plot indicates that there are no significant drops in the total within-cluster sum of squares after 4 clusters, so 4 clusters was used for our $k$-means analysis. 

### Visualizations

Four plots were generated to visualize the differences between clusters and popularity scores. Click through the tabset panels to explore the different visualizations.   

::: {.panel-tabset}

### Danceability, Energy, Liveness

```{r}
#| label: k-means analysis plot 1
#| warning: false
#| out-width: 100%

spotify_k4 |>
  select(Popularity, Danceability, Energy, Liveness, Cluster) |>
  ggpairs(aes(color = Cluster))
```

Songs in cluster 3 and 4 tended to have a higher popularity rating, while those in cluster 1 and 2 had the lowest, although there were no significant differences between the four clusters. All four clusters had relatively high danceability scores, with cluster 3 having the highest and cluster 2 having the lowest. Cluster 4 had the lowest energy score, while cluster 3 had the second lowest. Clusters 1 and 2 had similarly high energy scores. All four clusters had low liveness scores. 

### Key, Tempo, Duration

```{r}
#| label: k-means analysis plot 2
#| warning: false
#| out-width: 100%

spotify_k4 |>
  select(Popularity, Key, Tempo, Duration, Cluster) |>
  ggpairs(aes(color = Cluster))
```

### Speechiness, Acousticness, Instrumentalness

```{r}
#| label: k-means analysis plot 3
#| warning: false
#| out-width: 100%

spotify_k4 |>
  select(Popularity, Speechiness, Acousticness, Instrumentalness, Cluster) |>
  ggpairs(aes(color = Cluster))
```

### Valence, Mode, Loudness

```{r}
#| label: k-means analysis plot 4
#| warning: false
#| out-width: 100%

spotify_k4 |>
  select(Popularity, Valence, Mode, Loudness, Cluster) |>
  ggpairs(aes(color = Cluster))
```

:::