---
title: "Sentiment Analysis"
author: "Ben Snyderman"
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
library(tidyverse)
library(dplyr)
library(RColorBrewer)
library(wordcloud)


# Load datasets
load("data/sentiment-data.Rds")
```

### Sentiment Analysis

## What words are the most popular songs using?

*All data is from the top 100 spotify songs*


## Total Word Count

::: {.panel-tabset}

### Word Cloud

This visual shows a word cloud including the top words from all songs.


```{r}

#| fig.width = 20
#| fig.height = 16

set.seed(26)
my_palette2 <- brewer.pal(8, "Dark2")

word_counts_bleeped |>
with(wordcloud(words = word,
freq = n,
max.words = 100,
random.order = TRUE,
scale = c(5, 0.5),
rot.per = 0.15,
colors = my_palette2,
family = "arial"))


```



### List

This visual shows the top words used among all songs in general.

```{r}

ggplot(data=word_counts_bleeped_sliced, aes(x = reorder(word,frequency), y = frequency)) +
geom_col(fill = "dodgerblue2", show.legend = FALSE) +
labs(x = NULL, y = "Frequency") +
coord_flip()


```


:::




This visual shows the top words in terms of usage under each category of sentiment defined by nrc.

```{r}

ggplot(data=playlist_nrc_bleeped, aes(x = reorder(word,frequency), y = frequency)) +
geom_col(fill = "dodgerblue2", show.legend = FALSE) +
labs(x = NULL, y = "Frequency") +
facet_wrap(~ sentiment, ncol = 2, scales = "free") +
coord_flip()


```








## All Song Sentiments

::: {.panel-tabset}


### Top 10 Songs

This visual shows the top ten songs in terms of sentiment according to afinn.

```{r}


ggplot(afinn_best, aes(x = reorder(track_name, sentiment_score), y = sentiment_score)) +
  geom_col(fill = "chartreuse4", show.legend = FALSE) +
  labs(x = NULL, y = "Sentiment Score") +
coord_flip()
```


### Bottom 10 Songs

This visual shows the bottom ten songs in terms of sentiment according to afinn.

```{r}
ggplot(afinn_worst, aes(x = reorder(track_name, sentiment_score), y = sentiment_score)) +
  geom_col(fill = "red3", show.legend = FALSE) +
  labs(x = NULL, y = "Sentiment Score") +
coord_flip()
```


:::




## Pop vs. Rap

::: {.panel-tabset}

### Top 10 Songs

The next visual shows the top songs in terms of sentiment according to afinn categorized by genre.

```{r}

ggplot(data=afinn_rap_pop_good, aes(x = reorder(track_name,sentiment_score), y = sentiment_score)) +
geom_col(aes(fill = sentiment_score > 0), show.legend = FALSE) +
labs(x = NULL, y = "Sentiment Score") +
facet_wrap(~ playlist_genre, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = c("TRUE" = "chartreuse4", "FALSE" = "red3")) +
coord_flip()
```


### Bottom 10 Songs


The next visual shows the bottom songs in terms of sentiment according to afinn categorized by genre.

```{r}

ggplot(data=afinn_rap_pop_bad, aes(x = reorder(track_name,sentiment_score), y = sentiment_score)) +
geom_col(fill = "red3", show.legend = FALSE) +
labs(x = NULL, y = "Sentiment Score") +
facet_wrap(~ playlist_genre, ncol = 2, scales = "free_y") +
coord_flip()
```


:::

